Comparing R packages for calculation of persistent homology

### 2020-01-26 tdabench-paper/tdabench.Rmd
Added Abstract

---
title: Comparing R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  An abstract of less than 150 words.
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

>>>>

---
title: Comparing R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  With the increasing popularity of topological data analysis, several persistent homology software libraries have been implemented in R. The Dionysus and GUDHI C++ libraries have been implemented in the TDA package, and the Ripser C++ library has been implemented in the TDAstats package. All libraries feature Rips complex functionality. Additionally, within the TDA GUDHI package, the Alpha complex is available. These software represent powerful tools for analyzing the underlying topological structure of data. However, they are limited by computational restraints. High dimensional features and large point cloud sizes are limiting factors for persistent homology calculations. These limitations and guidelines for which persistent homology libraries to use have not been specified in R. To fill this gap, we benchmarked these libraries on a High-Powered Cluster. We benchmarked all Rips complex libraries for runtime. In addition, we compared Alpha and Rips for memory burden and runtime. We find that Alpha complexes run significantly faster than all Rips complex libraries for all data sets where data dimension is less than or equal to three. Alpha complexes were not computable on our systems for greater than 3 dimensions. For Rips complex calculations, run time performance was fastest for TDAstats and slowest for Dionysus. This difference was markedly clear at high dimensional features and point cloud values. From a computational efficiency perspective, if data dimension is less than or equal to 3, we recommend performing persistent homology with GUDHI Alpha complexes. If data dimension is higher, we recommend using TDAstats Rips Complex. 
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---
